---
title: "Building an Instructional Simulation App in Shiny"
author: "Homer White"
date: "June 17, 2015"
output: html_document
runtime: shiny
---

## Introdution and Prerequisites

Shiny makes it easy for R users to develop responsive, R-powered web applications.  As you probably know, either from your own intial rxplorations with Shiny or from the [Shiny Tutorial](http://shiny.rstudio.com/tutorial/), creating simple apps is no problem, and probably you have some ideas for 

This tutorial assumes that you have:

* **familiarity with R**.  We'll assume some basic facilty in R programming and that you can at least read and understand R code that creates custom plots in the base graphics sytem R system.  For the most part our explanation of R code will be limited to its relationship to app-building.
* **an introductory knowledge of Shiny**.  All necessary prerequisites inthis area can be acquired by watching Garret Grolemund's excellent three-part webinar on *How to Stat with Shiny*.  (See R Studio's [webinars](http://www.rstudio.com/resources/webinars/) page.)

## Version 1


Here is the first version:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_01/",
  options=list(
    width="100%", height=550
  )
)
```

## Version 2

Here is the second verison:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_02/",
  options=list(
    width="100%", height=550
  )
)
```


## Version 3

Here is the third version:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_03/",
  options=list(
    width="100%", height=550
  )
)
```

## Version 4

Here is the fourth version:

```{r echo = FALSE}
shinyAppDir(
  "example_sim_04/",
  options=list(
    width="100%", height=650
  )
)
```

## Version 5

```{r echo = FALSE}
shinyAppDir(
  "example_sim_05/",
  options=list(
    width="100%", height=800
  )
)
```

## Version 6

```{r echo = FALSE}
shinyAppDir(
  "example_sim_06/",
  options=list(
    width="100%", height=800
  )
)
```


## Version 7

```{r echo = FALSE}
shinyAppDir(
  "example_sim_07/",
  options=list(
    width="100%", height=800
  )
)
```

You may also view the final version on the <a href = "http://homer.shinyapps.io/example_sim_07" target = "_blank">web</a>.

Hope you enjoy!

