---
title: "Building an Instructional Simulation App in Shiny"
author: "Homer White"
date: "June 17, 2015"
output: html_document
runtime: shiny
---

## Introdution and Prerequisites

Shiny makes it easy for R users to develop responsive, R-powered web applications.  As you probably know, either from your own intial rxplorations with Shiny or from the [Shiny Tutorial](http://shiny.rstudio.com/tutorial/), creating simple apps is no problem, and probably you have some ideas for teaching apps tha toculd be written using just the tools developed in the Tutorial.

But some teaching apps appear quit ecomplex.  Consider, for example, [this app](http://homer.shinyapps.io/SlowGoodness) that aims to introduce the student to the $\chi^2$ Test for Goodness of Fit.  The app takes the user through a simulation process, keeping track of the results of simulations as they accumulate, permittnig the viewer to consider the results from several points of view, and allowing the viewer to start over, perhaps with new data.

This tutorial assumes that you have:

* **familiarity with R**.  We'll assume some basic facilty in R programming and that you can at least read and understand R code that creates custom plots in the base graphics sytem R system.  For the most part our explanation of R code will be limited to its relationship to app-building.
* **an introductory knowledge of Shiny**.  All necessary prerequisites inthis area can be acquired by watching Garret Grolemund's excellent three-part webinar on *How to Stat with Shiny*.  (See R Studio's [webinars](http://www.rstudio.com/resources/webinars/) page.)

## Version 1


Here is the first version:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_01/",
  options=list(
    width="100%", height=550
  )
)
```

## Version 2

Here is the second verison:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_02/",
  options=list(
    width="100%", height=550
  )
)
```


## Version 3

Here is the third version:

```{r, echo=FALSE}
shinyAppDir(
  "example_sim_03/",
  options=list(
    width="100%", height=550
  )
)
```

## Version 4

Here is the fourth version:

```{r echo = FALSE}
shinyAppDir(
  "example_sim_04/",
  options=list(
    width="100%", height=650
  )
)
```

## Version 5

```{r echo = FALSE}
shinyAppDir(
  "example_sim_05/",
  options=list(
    width="100%", height=800
  )
)
```

## Version 6

```{r echo = FALSE}
shinyAppDir(
  "example_sim_06/",
  options=list(
    width="100%", height=800
  )
)
```


## Version 7

```{r echo = FALSE}
shinyAppDir(
  "example_sim_07/",
  options=list(
    width="100%", height=800
  )
)
```

You may also view the final version on the <a href = "http://homer.shinyapps.io/example_sim_07" target = "_blank">web</a>.

Hope you enjoy!

